---
title: "Heart Disease Classification Pipeline with funcml"
subtitle: "From Formula Definition to Exported Results"
author: "Imad El Badisy"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    fig-pos: "H"
    geometry:
      - margin=1in
execute:
  echo: true
  warning: false
  message: false
  error: false
  cache: false
---

```{r setup}
library(funcml)

results_dir <- "heartdisease_pipeline_results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

set.seed(42)
```

# Objective

This document builds a complete binary classification pipeline using the internal
`heartdisease` dataset shipped with **funcml**. The target is `heart_disease`
(`"No"` / `"Yes"`), and the workflow proceeds from an explicit model formula to
model fitting, resampling-based evaluation, tuning, comparison, interpretation,
calibration, prediction, and exported results.

# Data

## Load and Prepare

```{r data-prep}
data_heart <- heartdisease

data_heart$sex <- factor(data_heart$sex)
data_heart$chest_pain <- factor(data_heart$chest_pain)
data_heart$blood_sugar <- factor(
  data_heart$blood_sugar,
  levels = c(FALSE, TRUE),
  labels = c("normal", "high")
)
data_heart$exercise_induced_angina <- factor(data_heart$exercise_induced_angina)
data_heart$heart_disease <- factor(
  data_heart$heart_disease,
  levels = c("No", "Yes")
)

data_cls <- data_heart
str(data_cls)
table(data_cls$heart_disease)
```

## Explicit Model Formula

```{r formula}
formula_cls <- heart_disease ~ age + sex + chest_pain + bp + cholesterol +
  blood_sugar + maximum_hr + exercise_induced_angina

formula_cls
```

# Baseline Model

We start with a random forest classifier because it is a strong tabular baseline
and supports probability prediction.

```{r baseline-fit}
fit_baseline <- fit(
  formula = formula_cls,
  data    = data_cls,
  model   = "ranger",
  spec    = list(num.trees = 300, mtry = 3, seed = 42)
)

fit_baseline
```

```{r baseline-predictions}
pred_class <- predict(fit_baseline, newdata = data_cls)
pred_prob  <- predict(fit_baseline, newdata = data_cls, type = "prob")

predictions <- data.frame(
  id = seq_len(nrow(data_cls)),
  truth = data_cls$heart_disease,
  predicted_class = pred_class,
  prob_no = pred_prob[, "No"],
  prob_yes = pred_prob[, "Yes"]
)

head(predictions)
```

# Resampling-Based Evaluation

```{r evaluation}
eval_baseline <- evaluate(
  data       = data_cls,
  formula    = formula_cls,
  model      = "ranger",
  resampling = cv(v = 5, seed = 42),
  metrics    = c("accuracy", "auc", "logloss"),
  type       = "prob",
  seed       = 42
)

eval_baseline
eval_baseline$summary
```

```{r evaluation-plot, fig.cap="Cross-validated performance for the baseline random forest classifier."}
plot(eval_baseline)
```

# Hyperparameter Tuning

```{r tuning}
grid_ranger <- expand.grid(
  num.trees = c(100, 200, 300),
  mtry      = c(2, 3, 5)
)

tuned_ranger <- tune(
  data       = data_cls,
  formula    = formula_cls,
  model      = "ranger",
  grid       = grid_ranger,
  resampling = cv(v = 5, seed = 42),
  metric     = "auc",
  type       = "prob",
  search     = "grid",
  seed       = 42
)

tuned_ranger$best
```

```{r tuning-plot, fig.cap="Ranger tuning results ranked by cross-validated AUC."}
plot(tuned_ranger)
```

# Final Model

```{r final-model}
fit_final <- fit(
  formula = formula_cls,
  data    = data_cls,
  model   = "ranger",
  spec    = as.list(tuned_ranger$best[1, c("num.trees", "mtry")])
)

fit_final
```

```{r final-predictions}
final_class <- predict(fit_final, newdata = data_cls)
final_prob  <- predict(fit_final, newdata = data_cls, type = "prob")

final_predictions <- data.frame(
  id = seq_len(nrow(data_cls)),
  truth = data_cls$heart_disease,
  predicted_class = final_class,
  prob_no = final_prob[, "No"],
  prob_yes = final_prob[, "Yes"]
)

head(final_predictions)
```

# Learner Comparison

```{r comparison}
cmp_cls <- compare_learners(
  data       = data_cls,
  formula    = formula_cls,
  models     = c("glm", "rpart", "ranger"),
  resampling = cv(v = 5, seed = 42),
  metrics    = c("accuracy", "auc", "logloss"),
  type       = "prob",
  seed       = 42
)

cmp_cls
cmp_cls$results
```

```{r comparison-plot, fig.cap="Comparison of GLM, CART, and random forest on the same folds."}
plot(cmp_cls)
```

# Interpretation

## Variable Importance

```{r importance}
vip_obj <- interpret(
  fit_final,
  data_cls,
  method = "vip"
)

vip_obj$result
```

```{r importance-plot, fig.cap="Model-based variable importance for the final random forest."}
plot(vip_obj)
```

## Permutation Importance

```{r permutation}
perm_obj <- interpret(
  fit_final,
  data_cls,
  method = "permute",
  metric = "auc",
  type   = "prob",
  nsim   = 10,
  seed   = 42
)

perm_obj$result
```

```{r permutation-plot, fig.cap="Permutation importance using AUC as the performance metric."}
plot(perm_obj)
```

## Calibration

```{r calibration}
cal_obj <- interpret(
  fit_final,
  data_cls,
  method    = "calibration",
  type      = "prob",
  pos_level = "Yes",
  bins      = 8
)

cal_obj$result
```

```{r calibration-plot, fig.cap="Calibration curve for predicted heart disease probability."}
plot(cal_obj, style = "curve")
```

# Optional SHAP Code

SHAP can be informative, but it is intentionally not evaluated here because it is
computationally expensive during document rendering.

```{r shap-optional, eval=FALSE}
high_risk <- data_cls[which.max(final_prob[, "Yes"]), , drop = FALSE]

shap_hr <- interpret(
  fit_final,
  data_cls,
  method  = "shap",
  newdata = high_risk,
  type    = "prob",
  nsim    = 50,
  seed    = 42
)

plot(shap_hr, kind = "waterfall")
```

# Export Results

The following chunk writes the dataset, formula, fitted objects, tabular outputs,
predictions, plots, and session information to `heartdisease_pipeline_results/`.

```{r export-results}
write_export <- function(x, path) {
  tab <- tryCatch(
    {
      if (!is.data.frame(x)) {
        x <- as.data.frame(x)
      }
      x[] <- lapply(x, function(col) {
        if (is.list(col)) {
          vapply(col, function(value) {
            paste(capture.output(str(value, max.level = 1)), collapse = " ")
          }, character(1))
        } else {
          col
        }
      })
      x
    },
    error = function(e) {
      data.frame(output = capture.output(print(x)))
    }
  )

  write.csv(tab, path, row.names = FALSE)
}

write.csv(data_cls, file.path(results_dir, "heartdisease_classification_data.csv"),
          row.names = FALSE)
writeLines(deparse(formula_cls), file.path(results_dir, "model_formula.txt"))

write.csv(predictions, file.path(results_dir, "baseline_predictions.csv"),
          row.names = FALSE)
write.csv(final_predictions, file.path(results_dir, "final_predictions.csv"),
          row.names = FALSE)

write_export(eval_baseline$summary,
             file.path(results_dir, "baseline_evaluation_summary.csv"))
write_export(tuned_ranger$results,
             file.path(results_dir, "tuning_results.csv"))
write_export(tuned_ranger$best,
             file.path(results_dir, "best_hyperparameters.csv"))
write_export(cmp_cls$results,
             file.path(results_dir, "learner_comparison.csv"))
write_export(vip_obj$result,
             file.path(results_dir, "variable_importance.csv"))
write_export(perm_obj$result,
             file.path(results_dir, "permutation_importance.csv"))
write_export(cal_obj$result,
             file.path(results_dir, "calibration_table.csv"))

saveRDS(fit_baseline, file.path(results_dir, "baseline_fit.rds"))
saveRDS(eval_baseline, file.path(results_dir, "baseline_evaluation.rds"))
saveRDS(tuned_ranger, file.path(results_dir, "tuned_ranger.rds"))
saveRDS(fit_final, file.path(results_dir, "final_fit.rds"))
saveRDS(cmp_cls, file.path(results_dir, "learner_comparison.rds"))
saveRDS(vip_obj, file.path(results_dir, "variable_importance.rds"))
saveRDS(perm_obj, file.path(results_dir, "permutation_importance.rds"))
saveRDS(cal_obj, file.path(results_dir, "calibration.rds"))

pdf(file.path(results_dir, "baseline_evaluation_plot.pdf"), width = 7, height = 5)
print(plot(eval_baseline))
dev.off()

pdf(file.path(results_dir, "tuning_plot.pdf"), width = 7, height = 5)
print(plot(tuned_ranger))
dev.off()

pdf(file.path(results_dir, "learner_comparison_plot.pdf"), width = 7, height = 5)
print(plot(cmp_cls))
dev.off()

pdf(file.path(results_dir, "variable_importance_plot.pdf"), width = 7, height = 5)
print(plot(vip_obj))
dev.off()

pdf(file.path(results_dir, "permutation_importance_plot.pdf"), width = 7, height = 5)
print(plot(perm_obj))
dev.off()

pdf(file.path(results_dir, "calibration_plot.pdf"), width = 7, height = 5)
print(plot(cal_obj, style = "curve"))
dev.off()

writeLines(capture.output(sessionInfo()),
           file.path(results_dir, "session_info.txt"))

sort(list.files(results_dir))
```

# Reuse

The exported `.rds` files preserve the fitted **funcml** objects for later R
analysis. The `.csv` files are plain tabular outputs for reporting, auditing, or
sharing outside R. The plot PDFs can be inserted directly into reports or
presentations.
